### 2021-01-01 paper/localpsf.tex
H matrix block cluster tree writing

In order to use the product-convolution approximation $\widetilde{\mathcal{A}}$, we convert $\widetilde{\mathcal{A}}$ to hierarchical matrix (H-matrix) format. Many classes of hierarchical matrices exist---H, uniform H1, H2, HODLR, HSS, HBS, and more. We use H1 matrices in our numerical results, but any of the other H-matrix formats could be used instead. The basic ideas behind H-matrices are as follows:
\begin{itemize} 
	\item Often, large dense matrices of practical interest may be permuted, then partitioned into blocks recursively, in such a way that many off-diagonal blocks of the matrix are numerically low rank, even if the matrix is high rank.
	\item Recursive algorithms can take advantage of this low rank block structure to perform matrix arithmetic fast. Conventional algorithms for matrix inversion, matrix factorization, matrix-vector products, matrix-matrix products, and matrix addition require either $O(n^2)$ or $O(n^3)$ time and memory, while the aforementioned recursive algorithms can perform these matrix operations in $O(n \log^a n)$ time and memory for hierarchical matrices (here $a=0, 1, 2$, or $3$ depending on the operation and type of H-matrix).
\end{itemize}
For more details on H-matrices, we recommend [CITES].

We use the standard geometrical clustering/adaptive cross H1 matrix construction method implemented within the HLibPro software package [CITE] to convert $\widetilde{\mathcal{A}}$ into H1 matrix format. We now summarize this method.

\paragraph{Cluster trees} First, the degrees of freedom associated with the columns of the matrix are hierarchically clustered into a \emph{cluster tree} using recursive hyperplane splitting. Since we use finite elements to discretize the problem, we associate the $i^\text{th}$ column of $\widetilde{\mathcal{A}}$ with the nodal point in $\mathbb{R}^d$ associated with the $i^\text{th}$ finite element basis function. The columns of the matrix thus correspond to a point cloud in $\mathbb{R}^d$. We split this point cloud into two roughly equally sized \emph{child} point clouds, using a hyperplane which is perpendicular to the coordinate axis direction in which the point cloud is widest (e.g., either the $x$, $y$, or $z$ axis in $3$D). The two child point clouds are split in the same way. This splitting process repeats recursively until the point clouds have less than a preset number of points (we use $32$ points). This hierarchical partitioning of the point cloud into smaller and smaller point clouds corresponds to a hierarchical partitioning of the columns of the matrix into smaller and smaller \emph{clusters} of columns, which we call a cluster tree. A depth-first search ordering of the cluster tree is then generated. When the columns of the matrix are permuted into this ordering, the columns associated with each cluster in the cluster tree are contiguous.

@@

\paragraph{Block cluster tree} Second, we partition the matrix (after reordering the rows and columns via the depth-first ordering described above) into a recursive hierarchy of mostly low rank blocks called the \emph{block cluster tree}.
The idea is that the interaction between far-away groups of degrees of freedom may be characterized by a small number of bulk properties of those groups.

>>>>

In order to use the product-convolution approximation $\widetilde{\mathcal{A}}$, we convert $\widetilde{\mathcal{A}}$ to hierarchical matrix (H-matrix) format. Many classes of hierarchical matrices exist---H1, uniform H1, H2, HODLR, HSS, HBS, and more. We use H1 matrices in our numerical results, but any of the other H-matrix formats could be used instead. The basic ideas behind H-matrices are as follows:
\begin{itemize} 
	\item Often, large dense matrices of practical interest may be permuted, then partitioned into blocks recursively, in such a way that many off-diagonal blocks of the matrix are numerically low rank, even if the matrix is high rank.
	\item Recursive algorithms can take advantage of this low rank block structure to perform matrix arithmetic fast. Conventional algorithms for matrix inversion, matrix factorization, matrix-vector products, matrix-matrix products, and matrix addition require either $O(n^2)$ or $O(n^3)$ time and memory, while the aforementioned recursive algorithms can perform these matrix operations in $O(n \log^a n)$ time and memory for hierarchical matrices (here $a=0, 1, 2$, or $3$ depending on the operation and type of H-matrix).
\end{itemize}
For more details on H-matrices, we recommend [CITES].

We convert $\widetilde{\mathcal{A}}$ into H1 matrix format using the standard geometrical clustering/adaptive cross method implemented within the HLIBpro software package [CITE]. In the remainder of this section we describe this method in detail.

\paragraph{Cluster trees} First, the degrees of freedom associated with the columns of the matrix are hierarchically clustered into a \emph{cluster tree} using recursive hyperplane splitting. Since we use finite elements to discretize the problem, we associate the $i^\text{th}$ column of $\widetilde{\mathcal{A}}$ with the nodal point in $\mathbb{R}^d$ associated with the $i^\text{th}$ finite element basis function. The columns of the matrix thus correspond to a point cloud in $\mathbb{R}^d$. We split this point cloud into two equally sized \emph{child} point clouds, using a hyperplane which is perpendicular to the coordinate axis direction in which the point cloud is widest (e.g., either the $x$, $y$, or $z$ axis in $3$D). The two child point clouds are split in the same way. This splitting process repeats until the point clouds have less than a preset number of points (we use $32$ points). This hierarchical partitioning of the point cloud into smaller and smaller point clouds corresponds to a hierarchical partitioning of the columns of the matrix into smaller and smaller \emph{clusters} of columns. This hierarchical partitioning of the columns forms a binary tree, which we call a cluster tree. The root of the tree is the set of all columns, and the leaves of the tree are clusters of columns that are not subdivided any further. A depth-first search ordering of the cluster tree leaves is then generated. When the columns of the matrix are permuted into this depth-first ordering, the columns associated with each cluster in the cluster tree are contiguous.

@@

\paragraph{Block cluster tree} Second, we partition the matrix into a recursive hierarchy of mostly low rank blocks called the \emph{block cluster tree}. The idea is that a block of the matrix is likely to be low rank if the point cloud associated with the rows of the block is far away from the point cloud associated with the columns of the block. This is reasonable to expect here because of the locality property of $\mathcal{A}$. Indeed, locality implies that many blocks of the matrix corresponding to far away point cloud clusters will be rank zero.

After reordering the rows and columns of $\widetilde{\mathcal{A}}$ via the depth-first ordering described above, we partition the reordered version of $\widetilde{\mathcal{A}}$ into a tree of $2 \times 2$ block matrices recursively. We use a geometric admissibility condition (discussed below) to decide which blocks to subdivide, and use the cluster trees for the rows and columns to determine how to subdivide those blocks. For the first stage of subdivision, let $r_1$ and $r_2$ be the children row clusters for the root of the row cluster tree, let $c_1$ and $c_2$ be the children column clusters for the root of the column cluster tree. The matrix $\widetilde{\mathcal{A}}$ is partitioned into blocks as follows:
\begin{equation*}
	\begin{bmatrix}
		\widetilde{\mathcal{A}}_{11} & \widetilde{\mathcal{A}}_{12} \\
		\widetilde{\mathcal{A}}_{21} & \widetilde{\mathcal{A}}_{22},
	\end{bmatrix}
\end{equation*}
where $\widetilde{\mathcal{A}}_{11}$ denotes the block of $\widetilde{\mathcal{A}}$ with rows $r_1$ and columns $c_1$, $\widetilde{\mathcal{A}}_{12}$ denotes the block of $\widetilde{\mathcal{A}}$ with rows $r_1$ and columns $c_2$, and similar for $\widetilde{\mathcal{A}}_{21}$ and $\widetilde{\mathcal{A}}_{22}$.

We now loop through the four blocks, $\widetilde{\mathcal{A}}_{11}$, $\widetilde{\mathcal{A}}_{12}$, $\widetilde{\mathcal{A}}_{21}$, and $\widetilde{\mathcal{A}}_{22}$, and decide which blocks should be subdivided further. For the purpose of explanation, consider $\widetilde{\mathcal{A}}_{12}$. If
\begin{equation}
	\label{eq:weak_admissibility_cond}
	\dist\left(r_1, c_2\right) \ge \eta \min\left(\diam\left(r_1\right), \diam\left(c_2\right)\right),
\end{equation}
then we mark $\widetilde{\mathcal{A}}_{12}$ as \emph{admissible} (expected to be low rank) and leave it alone. Here $\dist\left(r_1, c_2\right)$ is the Euclidean distance between the axis-aligned bounding box for the point cloud associated with the row cluster $r_1$, and the axis aligned bounding box for the point cloud associated with the column cluster $c_2$. The quantity $\diam\left(r_1\right)$ is the diameter of the axis aligned bounding box for the point cloud associated with the row cluster $r_1$, and $\diam\left(c_2\right)$ is the analogous diameter associated with the column cluster $c_2$. The quantity $\eta$ is a scalar constant; we use $\eta=2.0$. Basically, if the point clouds associated with $r_1$ and $c_2$ are far away from each other relative to their diameters, then we expect that the corresponding block of the matrix will be low rank. This process is repeated for the other blocks to determine which blocks are admissible and which are not. For us, the diagonal blocks $\widetilde{\mathcal{A}}_{11}$ and $\widetilde{\mathcal{A}}_{22}$ are not admissible because the distance between a point cloud and itself is zero.

Next, we subdivide all blocks that not admissible and are larger than a predetermined size (we use size $32 \times 32$), using the same process that was used to subdivide $\mathcal{\widetilde{A}}$. But now we subdivide a block based on the two child row clusters and two child column clusters for the rows and columns of that block. This subdivision process continues recursively until all blocks are either admissible, or smaller than the predetermined size mentioned above. The resulting hierarchical partitioning of matrix blocks forms a tree (quad-tree in $2$D, or oct-tree in $3$D), which is called the \emph{block cluster tree}. The root of the tree is the whole matrix, internal nodes in the tree are blocks that are subdivided, and the leaves of the tree are blocks that are either expected to be low rank, or are small.

Note that the cluster trees and block cluster tree are computed using only the geometric information about the degree of freedom locations, and do not depend on the values of the matrix entries of $\widetilde{\mathcal{A}}$.

\paragraph{Adaptive cross low rank approximation of blocks} WORK IN PROGRESS
